---
title: "Calibration curves are not only about linearity"
author: "Calibration simulation series"
format:
  html:
    toc: true
execute:
  echo: true
  warning: false
  message: false
---

# Calibration curves are not only about linearity

In the previous post we compared different ways to distribute a fixed number of calibration measurements.

The conclusion was simple:

A calibration design does not only define the curve.

It defines what information we can extract from the experiment.

But there is another problem that is often ignored:

**Are all calibration points equally precise?**


# The hidden assumption behind ordinary least squares

Ordinary least squares assumes that the variability of measurements is constant.

In other words:

small concentrations and high concentrations should have comparable uncertainty.

But many analytical methods do not behave this way.

Examples:

- instrumental noise
- low level quantification
- detector response
- trace analysis

often show increasing variability across the calibration range.

This phenomenon is called **heteroscedasticity**.

# Simulation setup

We compare the same three calibration designs:

- 15 different calibration levels
- 5 levels measured in triplicate
- 3 levels measured five times

The total number of measurements is always 15.

```{r}
#| label: resources
library(here)
library(data.table)
library(ggplot2)

source(here("R", "response_functions.R"))
source(here("R", "variance_functions.R"))
source(here("R", "design_functions.R"))
source(here("R", "simulation_functions.R"))
source(here("R", "fit_functions.R"))
source(here("R", "prediction_functions.R"))
source(here("R", "metrics_functions.R"))
source(here("R", "plotting_functions.R"))
```


```{r}
#| label: set_designs
set.seed(123)

mean_fun <- f_linear

variance_fun <- function(x) {

  sigma_constant_rsd(
    x,
    mean_fun = mean_fun,
    rsd = 0.10
  )

}

designs <- list(
  "15 x 1" = design_15x1(1, 100),
  "5 x 3"  = design_5x3(1, 100),
  "3 x 5"  = design_3x5(1, 100)
)
```

## Simulate calibration data

```{r}
#| label: calibration_data
calibration_data <- lapply(
  designs,
  function(design) {

    simulate_dataset(
      design,
      mean_fun,
      variance_fun
    )

  }
)

plot_design(
  calibration_data[["15 x 1"]],
  mean_fun
)
```

The underlying response is still perfectly linear, the difference is only in the variance.

## Visualising heteroscedasticity

A useful diagnostic is the residual.

If variability increases with concentration, residuals should spread out.

```{r}
#| label: residuals
model_example <- fit_ols(
  calibration_data[["15 x 1"]]
)

plot_residuals(
  model_example
)
```

## Fit ordinary least squares models

For now, all designs are analysed using the same model.

```{r}
#| label: fit_models 
models <- lapply(
  calibration_data,
  fit_ols
)
```

## Generate new unknown samples

A calibration curve should not only describe the standards used to build it.

It should predict future samples.

```{r}
#| label: test_models
test_data <- generate_test_set(
  n = 1000,
  xmin = 0,
  xmax = 100,
  mean_fun = mean_fun,
  variance_fun = variance_fun
)
```

## Prediction performance

Convert model output into a common format.

```{r}
#| label: prediction_summary
prediction_tables <- lapply(
  models,
  predict_calibration_model,
  test_data = test_data
)
```

## Prediction error

```{r}
#| label: prediction_error
lapply(
  prediction_tables,
  metric_rmse
)
```

The result is a single number, but it hides where the model performs poorly.

Let's look at the full profile.

```{r}
#| label: fig_prediction_error
plot_prediction_error(
  prediction_tables[["15 x 1"]]
)

plot_prediction_error(
  prediction_tables[["5 x 3"]]
)

plot_prediction_error(
  prediction_tables[["3 x 5"]]
)
```

## Prediction uncertainty

Another important question:

How wide is the prediction interval?

```{r}
#| label: prediction_uncertainty
lapply(
  prediction_tables,
  metric_mean_pi_width
)
```

The calibration curve may look acceptable while uncertainty is poorly estimated.

## Can we detect heteroscedasticity?

With no replicates, estimating variability at each concentration is difficult.

Let's test whether the residual pattern is detected.

```{r}
#| label: heteroscedasticity_test
hetero_detected <- sapply(
  models,
  metric_hetero_detected
)

hetero_detected
```

## Conclusions

When the variance is not constant:

- the response can still be linear
- the calibration curve can still look good
- uncertainty estimation can be wrong

Replicates provide information about variability, but there is a trade-off: more replicates mean fewer concentration levels.

## Next question

If we know that variability changes with concentration: should we still use ordinary least squares?

This leads to weighted least squares.
