---
title: "How should we spend 15 calibration measurements?"
author: "Calibration simulation series"
format:
  html:
    toc: true
execute:
  echo: true
  warning: false
  message: false
---

# How should we spend 15 calibration measurements?

Analytical methods often require a calibration curve built from several standards.

A common recommendation is:

- use at least 5–7 calibration levels
- distribute them across the working range
- evaluate linearity

But a practical question is rarely discussed:

**If the total number of measurements is fixed, where should we spend them?**

Imagine we have exactly 15 injections available.

Three possible designs:

1. 15 different calibration levels
2. 5 calibration levels measured in triplicate
3. 3 calibration levels measured five times

Which one gives us the most useful information?

## A provocative idea

A chemometrician once proposed:

> If the response is linear, use two extreme points, replicate them many times, and use a central point to check linearity.

At first this sounds wrong compared with common laboratory practice.

But it raises a more fundamental question:

**Are we trying to estimate the shape of the curve, or the uncertainty of the measurement?**

These are different problems.

## Simulation setup

We will compare three calibration designs.

```{r}
#| label: resources
library(here)
library(data.table)
library(ggplot2)

source(here("R", "response_functions.R"))
source(here("R", "variance_functions.R"))
source(here("R", "design_functions.R"))
source(here("R", "simulation_functions.R"))
source(here("R", "fit_functions.R"))
source(here("R", "prediction_functions.R"))
source(here("R", "metrics_functions.R"))
source(here("R", "plotting_functions.R"))
```

We start with a simple linear response and constant variance.

```{r}
#| label: set_designs
set.seed(123)

mean_fun <- f_linear

variance_fun <- sigma_homo

designs <- list(
  "15 x 1" = design_15x1(1, 100),
  "5 x 3"  = design_5x3(1, 100),
  "3 x 5"  = design_3x5(1, 100)
)
```

## Visual comparison of designs

```{r}
#| label: compare_designs
design_plot_data <- rbindlist(
  lapply(
    names(designs),
    function(name) {

      dt <- copy(designs[[name]])
      dt[, design := name]
      dt

    }
  )
)

plot_design_comparison(design_plot_data)
```

The three approaches use the same number of measurements.

The difference is where the information is allocated.

## Simulate calibration data

```{r}
#| label: calibration_data
calibration_data <- lapply(
  designs,
  function(design) {

    simulate_dataset(
      design,
      mean_fun,
      variance_fun
    )

  }
)

plot_design(
  calibration_data[["15 x 1"]],
  mean_fun
)
```

## Fit a classical linear model

For this first example we use ordinary least squares.

```{r}
#| label: fit_models 
models <- lapply(
  calibration_data,
  fit_ols
)
```

## Generate new unknown samples

A calibration curve should not only describe the standards used to build it.

It should predict future samples.

```{r}
#| label: test_models
test_data <- generate_test_set(
  n = 1000,
  xmin = 0,
  xmax = 100,
  mean_fun = mean_fun,
  variance_fun = variance_fun
)
```

## Prediction performance

Convert model output into a common format.

```{r}
#| label: prediction_summary
prediction_tables <- lapply(
  models,
  predict_calibration_model,
  test_data = test_data
)
```

## Prediction error

```{r}
#| label: prediction_error
lapply(
  prediction_tables,
  metric_rmse
)
```

The result is a single number, but it hides where the model performs poorly.

Let's look at the full profile.

```{r}
#| label: fig_prediction_error
plot_prediction_error(
  prediction_tables[["15 x 1"]]
)

plot_prediction_error(
  prediction_tables[["5 x 3"]]
)

plot_prediction_error(
  prediction_tables[["3 x 5"]]
)
```

## Prediction uncertainty

Another important question:

How wide is the prediction interval?

```{r}
#| label: prediction_uncertainty
lapply(
  prediction_tables,
  metric_mean_pi_width
)
```

A method can have a good average error but still have poor uncertainty estimation.

## First conclusions

In this ideal scenario:

- the response is perfectly linear
- the variance is constant
- the model assumptions are correct

The different designs can perform similarly.

But the situation changes when reality enters:

- variance changes with concentration
- the curve approaches a plateau
- low concentration points have higher relative variability

Those are the scenarios we will explore next.

---

## Repository

All simulations are reproducible.

The complete code is available in the repository:

- response functions
- variance models
- calibration designs
- regression models
- prediction metrics
- plots

The goal is not to find a single "best" calibration design.

The goal is to understand what information each design provides.