---
title: "How should we spend 15 calibration measurements?"
author: "Calibration simulation series"
format:
  html:
    toc: true
execute:
  echo: true
  warning: false
  message: false
---

# How should we spend 15 calibration measurements?

Calibration curves are among the most common tools in analytical chemistry.

We prepare standards with known concentration:

$$
x = \text{known concentration}
$$

we measure the instrument response:

$$
y = \text{instrument signal}
$$

and we build a model describing the relationship between concentration and response.

The final purpose of a calibration curve is not the curve itself.

The final purpose is:

**given a new measured signal, estimate the unknown concentration.**

This makes calibration an inverse problem.

Although the final objective is to estimate concentration, the calibration model is usually fitted in the opposite direction.

The reason is simple: calibration standards have known concentrations.

$$
y=f(x)
$$

and only during unknown sample analysis do we invert the model to obtain:

$$
x=f^{-1}(y)
$$

This distinction will become important later in the series.

# The usual approach

Most analytical guidelines recommend:

- several calibration levels
- a distribution across the working range
- evaluation of linearity

A common practical recommendation is:

- at least 5–7 calibration points
- no repeated levels

However, one practical question is rarely discussed:

**If the total number of measurements is fixed, where should we spend them?**

Imagine having exactly 15 injections available.

Possible strategies:

1. 15 different concentrations measured once
2. 5 concentrations measured in triplicate
3. 3 concentrations measured five times

All designs use the same number of measurements.

However, they do not provide the same information.

# A provocative idea

A chemometrician once suggested:

> If the response is linear, measure only the extreme concentrations repeatedly and use a central point to check linearity.

This sounds very different from common laboratory practice.

But it raises a fundamental question.

What information are we actually trying to obtain from calibration measurements?

Calibration points can be used to estimate:

- the shape of the response curve
- the uncertainty of the measurement

These are different objectives.

A design with many concentration levels is informative about the shape.

A design with many replicates is informative about variability.

The best strategy depends on which source of uncertainty dominates.

# Simulation philosophy

In a real laboratory experiment we only observe one calibration dataset.

We do not know the true concentration-response relationship.

Simulation allows us to create a controlled experiment.

We can know:

- the real concentration
- the real response
- the real uncertainty

and evaluate how well each calibration strategy recovers the concentration.

# Setup

```{r}
#| label: resources

library(here)
library(data.table)
library(ggplot2)

source(here("R", "design_functions.R"))
source(here("R", "fit_functions.R"))
source(here("R", "inversion_functions.R"))
source(here("R", "metrics_functions.R"))
source(here("R", "plotting_functions.R"))
source(here("R", "prediction_functions.R"))
source(here("R", "response_functions.R"))
source(here("R", "simulation_functions.R"))
source(here("R", "variance_functions.R"))
source(here("R", "weight_functions.R"))
```

# Calibration designs

We compare three ways of spending the 15 measurements.

```{r}
#| label: set_designs

set.seed(123)
mean_fun <- f_linear
variance_fun <- sigma_homo

designs <- list(
  "15 x 1" = design_15x1(1, 100),
  "5 x 3" = design_5x3(1, 100),
  "3 x 5" = design_3x5(1, 100)
)
```

# Where are the measurements?

```{r}
#| label: compare_designs

design_plot_data <- rbindlist(
  lapply(
    names(designs),
    function(name) {

      dt <- copy(designs[[name]])
      dt[, design := name]
      dt
    }
  )
)

plot_design_comparison(
  design_plot_data
)
```

The three strategies have the same number of measurements, only the distribution changes.

# Generate calibration data

```{r}
#| label: calibration_data

calibration_data <- lapply(
  designs,
  function(design) {

    simulate_dataset(
      design,
      mean_fun,
      variance_fun
    )

  }
)
```

```{r}
#| label: show_calibration

for (title in names(calibration_data)) {
  i <- calibration_data[[title]]
  print(
    plot_design(i, mean_fun) +
      labs(title = paste0("Design ", title))
  )
}
```

# Fit calibration models

The calibration model is built in the usual analytical direction.

We know the concentration of the standards, therefore we fit:

$$
y = f(x)
$$

```{r}
#| label: fit_models

models <- lapply(
  calibration_data,
  fit_ols
)
```

# Testing the calibration

A calibration curve is useful only if it predicts unknown samples.

We simulate unknown samples with known concentration.

```{r}
#| label: test_samples

test_data <- generate_test_set(
  n = 1000,
  xmin = 1,
  xmax = 100,
  mean_fun = mean_fun,
  variance_fun = variance_fun
)
```

# From signal back to concentration

For each simulated unknown:

1. we observe a signal
2. we apply the calibration model
3. we estimate the concentration

```{r}
#| label: inverse_prediction

prediction_tables <- lapply(
  models,
  predict_calibration_model,
  test_data = test_data
)
```

# How accurate are the concentrations?

The first question is:

**How close is the estimated concentration to the true concentration?**

A useful calibration should minimise the difference between predicted and real values.

```{r}
#| label: rmse

rmse_results <- sapply(
  prediction_tables,
  metric_rmse
)

rmse_results
```

The RMSE summarizes the average prediction error.

However, a single number can hide where the calibration performs poorly.

A model may work well in the middle of the range and become inaccurate near the limits.

```{r}
#| label: error_profile

prediction_profile <- lapply(
  prediction_tables,
  summarise_prediction_profile
)

for (title in names(prediction_tables)) {
  i <- prediction_tables[[title]]
  print(
    plot_prediction_error(i) +
      labs(title = paste0("Design ", title))
  )
}
```

The error is almost constant across the entire calibration range.
However, because the concentration range spans one or more orders of magnitude, the same absolute error corresponds 
to a much larger relative error at low concentrations than at high concentrations.

This is not surprising.

The model was fitted using Ordinary Least Squares (OLS), which assumes that the measurement errors 
have constant variance across the calibration range.
In other words, OLS gives the same importance to every calibration point, independently of its concentration.

OLS is often the default choice for laboratory calibration because it is simple, 
widely implemented, and works well when its assumptions are satisfied.

# Prediction intervals

A calibration model does not only provide a concentration estimate.

It should also provide an estimate of uncertainty.

A result such as:

> concentration = 50 mg/L

is incomplete.

A more realistic analytical result is:

> concentration = 50 mg/L, with an expected uncertainty

A prediction interval describes the range where a future measurement is expected to fall.

This is different from a confidence interval.

A confidence interval describes uncertainty in the estimated average response.

A prediction interval describes uncertainty for a future individual measurement.

For analytical chemistry, the second concept is usually the relevant one.

A future sample has its own measurement variability, and the calibration model must account for:

- uncertainty in the fitted calibration curve
- variability of future measurements

# Prediction interval coverage

A 95% prediction interval should have approximately 95% coverage.

This means that if we simulate many unknown samples:

- about 95% of the true concentrations should fall inside the calculated interval
- about 5% may fall outside

Coverage therefore tells us whether the uncertainty model is realistic.

The prediction interval is calculated in response space and then converted back to concentration.

```{r}
#| label: uncertainty

coverage_results <- sapply(
  prediction_tables,
  metric_coverage
)

coverage_results
```

A value close to 0.95 indicates a well calibrated uncertainty estimate.

Values much lower than 0.95 mean that intervals are too optimistic.

Values much higher than 0.95 mean that intervals may be unnecessarily wide.

# Interval width

Coverage alone is not enough.

A very wide interval can always achieve high coverage.

A useful calibration needs uncertainty that is both:

- reliable
- informative

Therefore we also examine the average prediction interval width.

```{r}
#| label: interval_width

width_results <- sapply(
  prediction_tables,
  metric_mean_pi_width
)

width_results
```

# First conclusions

In this first controlled scenario:

- the response is perfectly linear
- the variance is constant across the range
- the model assumptions are correct

The three designs perform similarly.

The design with 15 independent calibration levels gives the lowest RMSE in this simulation.

This makes sense.

When the true relationship is linear, spreading measurements across the whole range gives more information about the slope and intercept.

The replicated designs provide more information about variability, but less information about the shape of the calibration line.

However, this does not mean that 15 independent points are always the best choice.

The answer depends on where the uncertainty comes from.

If uncertainty is mainly caused by estimating the curve shape, additional levels are valuable.

If uncertainty is mainly caused by measurement variability, replicates become more useful.

The important question is therefore not:

**How many calibration points should we use?**

but:

**How should we distribute our measurements to obtain the information we need?**

# Next

So far we assumed that every concentration level had the same absolute variability.

However, many analytical methods behave differently.

The relative precision can remain approximately constant across the calibration range.

This means that the standard deviation increases with concentration, while the RSD remains similar.

When this happens, should every calibration point contribute equally to the regression?

Or should points with lower uncertainty have a larger influence?

This leads to weighted least squares.