---
title: "Weighted least squares: fixing the wrong problem?"
author: "Calibration simulation series"
format:
  html:
    toc: true
execute:
  echo: true
  warning: false
  message: false
---

# Weighted least squares: fixing the wrong problem?

In the previous post we introduced a common problem in analytical calibration:

**the variability of measurements may change across the calibration range.**

A linear calibration can still look correct.

But the uncertainty associated with low and high concentrations can be very different.

---

# The next step

If variability changes with concentration:

Should we still use ordinary least squares?

Ordinary least squares gives the same importance to every point.

But a point with high uncertainty should probably contribute less than a highly precise point.

This is the idea behind **weighted least squares (WLS).**

---

# The hidden challenge

Weighted regression requires weights.

But where do these weights come from?

Common choices include:

- experimentally estimated weights
- \(1/x\)
- \(1/x^2\)

Only the first option is statistically sound, the other two are common in analytical chemistry as they model common response uncertainty patterns.

Are they equivalent?

Let's simulate the problem.

# Simulation setup

We use the same heteroscedastic linear system from the previous episode.

The response is linear.

The variance increases with concentration.

The only difference is the regression model.

```{r}
#| label: resources
library(here)
library(data.table)
library(ggplot2)

source(here("R", "response_functions.R"))
source(here("R", "variance_functions.R"))
source(here("R", "design_functions.R"))
source(here("R", "simulation_functions.R"))
source(here("R", "fit_functions.R"))
source(here("R", "prediction_functions.R"))
source(here("R", "metrics_functions.R"))
source(here("R", "plotting_functions.R"))
source(here("R", "weight_functions.R"))
```

```{r}
#| label: set_designs
set.seed(123)

mean_fun <- f_linear

variance_fun <- function(x) {

  sigma_constant_rsd(
    x,
    mean_fun = mean_fun,
    rsd = 0.10
  )

}

designs <- list(
  "15 x 1" = design_15x1(1, 100),
  "5 x 3"  = design_5x3(1, 100),
  "3 x 5"  = design_3x5(1, 100)
)
```

## Simulate calibration data

Let's use the data from the previous episode:

```{r}
#| label: calibration_data
calibration_data <- lapply(
  designs,
  function(design) {

    simulate_dataset(
      design,
      mean_fun,
      variance_fun
    )

  }
)

plot_design(
  calibration_data[["15 x 1"]],
  mean_fun
)
```

## Fit different regression models

We compare:

- ordinary least squares
- WLS with experimentally estimated weights
- WLS with \[1/x\]
- WLS with \[1/x^2\]
- ideal WLS using the true variance (simulation benchmark)


```{r}
#| label: compare_models
models <- lapply(
  calibration_data,
  function(dt) {

    list(

      "OLS" = fit_ols(dt),

      "WLS estimated" = fit_wls_empirical(dt),

      "WLS 1/x" = fit_wls_x(dt),

      "WLS 1/x2" = fit_wls_x2(dt),

      "WLS true" = fit_wls_true(dt)

    )

  }
)
```

## Generate new samples

A calibration model should predict unknown samples.

```{r}
#| label: test_models
test_data <- generate_test_set(
  n = 1000,
  xmin = 0,
  xmax = 100,
  mean_fun = mean_fun,
  variance_fun = variance_fun
)
```

## Prediction performance

Convert model output into a common format.

```{r}
#| label: prediction_summary
prediction_tables <- lapply(
  models,
  function(model_set) {

    lapply(
      model_set,
      predict_calibration_model,
      test_data = test_data
    )

  }
)
```

## Prediction error

```{r}
#| label: prediction_error
rmse_results <- rbindlist(
  lapply(
    names(prediction_tables),
    function(design_name) {

      data.table(
        design = design_name,
        model = names(prediction_tables[[design_name]]),
        rmse = sapply(
          prediction_tables[[design_name]],
          metric_rmse
        )
      )

    }
  )
)

rmse_results
```

The result is a single number, but it hides where the model performs poorly.

Let's look at the full profile.


```{r}
#| label: fig_prediction_error
plot_prediction_error(
  prediction_tables[["5 x 3"]][["OLS"]]
)

plot_prediction_error(
  prediction_tables[["5 x 3"]][["WLS estimated"]]
)

plot_prediction_error(
  prediction_tables[["5 x 3"]][["WLS 1/x"]]
)

plot_prediction_error(
  prediction_tables[["5 x 3"]][["WLS 1/x2"]]
)

plot_prediction_error(
  prediction_tables[["5 x 3"]][["WLS true"]]
)

plot_prediction_error(
  prediction_tables[["3 x 5"]][["OLS"]]
)

plot_prediction_error(
  prediction_tables[["3 x 5"]][["WLS estimated"]]
)

plot_prediction_error(
  prediction_tables[["3 x 5"]][["WLS 1/x"]]
)

plot_prediction_error(
  prediction_tables[["3 x 5"]][["WLS 1/x2"]]
)

plot_prediction_error(
  prediction_tables[["3 x 5"]][["WLS true"]]
)
```

## Prediction uncertainty

Accuracy is not the only objective.

A calibration model should also estimate uncertainty correctly.

```{r}
#| label: prediction_uncertainty
pi_width_results <- summarise_metrics(
  prediction_tables,
  metric_mean_pi_width
)

pi_width_results
```

## Coverage of prediction intervals

A 95% prediction interval should contain the true value approximately 95% of the time.

```{r}
#| label: prediction_coverage
coverage_results <- summarise_metrics(
  prediction_tables,
  metric_coverage
)

coverage_results
```

## Comparing the residual structure

```{r}
ggplot(
  rmse_results,
  aes(
    x = model,
    y = rmse
  )
) +
geom_col() +
facet_wrap(~design)
```

## Conclusions

Weighted least squares can improve calibration when variability changes with concentration, but the choice of weights matters.

Using \[1/x^2\] is common, but it assumes a constant relative standard deviation (RSD) across the entire calibration range.

The real question is: do our weights describe the actual analytical uncertainty?

## Next

In the next episode we will test a more realistic situation: what if the RSD increases moving towards the limit of quantification (LoQ)?
