---
title: "When variability changes: introducing weighted least squares"
author: "Calibration simulation series"
format:
  html:
    toc: true
execute:
  echo: true
  warning: false
  message: false
---

# When variability changes: introducing weighted least squares

In the previous episode we compared different ways of distributing a fixed number of calibration measurements.

We assumed an ideal situation:

- the response was perfectly linear
- every concentration had the same absolute variability
- ordinary least squares provided an appropriate model

Under these conditions, spreading measurements across the calibration range worked well.

But real analytical methods are often different.

Many instrumental responses show approximately constant **relative precision** across the working range.

This means that the relative standard deviation (RSD) remains similar,
while the absolute variability increases with concentration.

# From constant variance to constant RSD

A constant RSD means: $\text{RSD} = \frac{\sigma}{x}$ remains approximately constant.

Therefore $\sigma$ increases as concentration increases.

A high concentration standard may show much larger absolute scatter than a low concentration standard.

This is not necessarily a problem.

It is simply a different variance structure.

The question is:

**Should all calibration points contribute equally to the regression?**

# The limitation of ordinary least squares

Ordinary least squares (OLS) estimates the calibration line by minimising the sum of squared residuals.

Every point receives the same importance.

This works well when all measurements have the same variance.

However, with constant RSD:

- low concentration points have smaller absolute uncertainty
- high concentration points have larger absolute uncertainty

Treating them equally may not be the most efficient choice.

A deviation of 5 response units at low concentration may represent a large relative error.

The same deviation at high concentration may be much less important.

# Weighted least squares

Weighted least squares (WLS) modifies the regression.

Instead of giving every observation the same influence, each point receives a weight.

The general principle is:

$$
w_i=\frac{1}{\sigma_i^2}
$$

where:

- $w_i$ is the weight
- $\sigma_i$ is the expected standard deviation

Points with lower uncertainty receive higher weight.

The difficulty is that in a real calibration we usually do not know the true variance.

Therefore:

**how should we choose the weights?**

# Different ways to choose weights

In this simulation we compare different strategies.

## True variance

Because we generated the data ourselves, we know the variance model.

This allows us to calculate the theoretical optimal weights.

This is not available in routine analysis, but it provides a useful reference.

## Empirical variance estimation

In practice, replicate measurements can be used to estimate variability.

The estimated standard deviation can then be converted into weights.

This approach is closer to a real laboratory workflow.

## Common analytical approximations

When the exact variance structure is unknown, simple weighting schemes are often used.

Two common choices are:

$$
w=\frac{1}{x}
$$

and:

$$
w=\frac{1}{x^2}
$$

These assume a specific relationship between concentration and variability.

The question is not only whether WLS improves the fit.

The important question is:

**how does the weighting strategy affect the final analytical result?**

# Simulation setup

We keep the same experimental question as episode 1.

We have 15 measurements available.

How should we distribute them?

```{r}
#| label: resources
library(here)
library(data.table)
library(ggplot2)

source(here("R", "design_functions.R"))
source(here("R", "fit_functions.R"))
source(here("R", "inversion_functions.R"))
source(here("R", "metrics_functions.R"))
source(here("R", "plotting_functions.R"))
source(here("R", "prediction_functions.R"))
source(here("R", "response_functions.R"))
source(here("R", "simulation_functions.R"))
source(here("R", "variance_functions.R"))
source(here("R", "weight_functions.R"))
```

```{r}
#| label: set_designs
set.seed(123)

mean_fun <- f_linear

variance_fun <- function(x) {

  sigma_constant_rsd(
    x,
    mean_fun = mean_fun,
    rsd = 0.10
  )

}

designs <- list(
  "15 x 1" = design_15x1(1, 100),
  "5 x 3"  = design_5x3(1, 100),
  "3 x 5"  = design_3x5(1, 100)
)
```

## Simulate calibration data

```{r}
#| label: calibration_data
calibration_data <- lapply(
  designs,
  function(design) {

    simulate_dataset(
      design,
      mean_fun,
      variance_fun
    )

  }
)
```

```{r}
#| label: show_calibration

for (title in names(calibration_data)) {
  i <- calibration_data[[title]]
  print(
    plot_design(i, mean_fun) +
      labs(title = paste0("Design ", title))
  )
}
```

The scatter increases with concentration.

The relative variability remains approximately constant.

## Fit different regression models

For each design we compare:

- OLS
- WLS using the true variance
- WLS using empirical variance
- WLS using $1/x$
- WLS using $1/x^2$


```{r}
#| label: compare_models
models <- lapply(
  calibration_data,
  function(dt) {

    list(
      "OLS" = fit_ols(dt),
      "WLS true" = fit_wls_true(dt),
      "WLS estimated" = fit_wls_empirical(dt),
      "WLS 1/x" = fit_wls_x(dt),
      "WLS 1/x2" = fit_wls_x2(dt)
    )

  }
)
```

# Testing unknown samples

The final goal is still the same: predict unknown concentrations.

```{r}
#| label: test_models
test_data <- generate_test_set(
  n = 1000,
  xmin = 1,
  xmax = 100,
  mean_fun = mean_fun,
  variance_fun = variance_fun
)
```

# Convert signals into concentrations

```{r}
#| label: prediction_summary
prediction_tables <- lapply(
  models,
  function(model_set) {

    lapply(
      model_set,
      predict_calibration_model,
      test_data = test_data
    )

  }
)
```

## Prediction error

```{r}
#| label: prediction_error
rmse_results <- rbindlist(
  lapply(
    names(prediction_tables),
    function(design_name) {

      data.table(
        design = design_name,
        model = names(prediction_tables[[design_name]]),
        rmse = sapply(
          prediction_tables[[design_name]],
          metric_rmse
        )
      )

    }
  )
)

rmse_results
```


```{r}
#| label: fig_prediction_error
for(design in names(prediction_tables)){

  for(model in names(prediction_tables[[design]])){

  print(
    plot_prediction_error(
      prediction_tables[[design]][[model]]
    ) +
    labs(title = paste(design, model))
    )
  }
}
```

# Prediction interval coverage

A good calibration should also estimate uncertainty correctly.

For a 95% prediction interval we expect approximately 95% coverage.

```{r}
#| label: coverage

coverage_results <- summarise_metrics(
  prediction_tables,
  metric_coverage
)

coverage_results
```

---

# Interval width

Coverage alone is not sufficient.

Intervals should be reliable but not unnecessarily wide.

```{r}
#| label: width

width_results <- summarise_metrics(
  prediction_tables,
  metric_mean_pi_width
)

width_results
```

---

# Conclusions

When variability changes with concentration, the assumption behind OLS is no longer satisfied.

Weighted least squares can improve calibration by accounting for different uncertainty levels across the range.

However, the weighting strategy matters.

Using the true variance gives the best possible benchmark, but it is not available in routine work.

Empirical weighting requires enough information to estimate variability.

Simple approaches such as $1/x$ or $1/x^2$ are convenient, but they rely on assumptions.

The important question is therefore not:

"Should we always use WLS?"

but:

"How well does the chosen weighting model describe the real variance structure?"

# Next

In the next episode we will explore this question by comparing weighting strategies under more realistic variance models.
