---
title: "Calibration curves are not only about linearity"
author: "Andrea Bazzano"
format:
  html:
    toc: true
execute:
  echo: true
  warning: false
  message: false
---

# The hidden assumption

A calibration curve can be perfectly linear and still be statistically uncomfortable. The reason is that OLS assumes more than a straight mean response. It also assumes that the response variability is approximately constant across the calibration range.

Many analytical methods do not behave that way. At low levels, signal processing and background noise may dominate. At higher levels, the absolute scatter can increase with response. The result is heteroscedasticity: the variance changes with concentration.

This episode keeps the response linear and changes only the measurement variability. That separation is useful for analysts because it shows that linearity and constant variance are different questions.

# Setup

```{r}
#| label: resources

library(here)
library(data.table)
library(ggplot2)
library(knitr)

source(here("R", "design_functions.R"))
source(here("R", "weight_functions.R"))
source(here("R", "fit_functions.R"))
source(here("R", "inversion_functions.R"))
source(here("R", "metrics_functions.R"))
source(here("R", "plotting_functions.R"))
source(here("R", "prediction_functions.R"))
source(here("R", "response_functions.R"))
source(here("R", "simulation_functions.R"))
source(here("R", "variance_functions.R"))
```

```{r}
#| label: setup

set.seed(123)

mean_fun <- f_linear
sigma_fun <- sigma_linear_constant_rsd

designs <- list(
  "15 x 1" = design_15x1(1, 100),
  "5 x 3" = design_5x3(1, 100),
  "3 x 5" = design_3x5(1, 100)
)
```

# The variance pattern

```{r}
#| label: variance-pattern
#| fig-cap: "Under constant RSD, absolute response standard deviation increases with concentration."

variance_grid <- data.table(
  x = seq(1, 100, length.out = 200)
)

variance_grid[, sigma := sigma_fun(x)]
variance_grid[, rsd := sigma / mean_fun(x)]

ggplot(variance_grid, aes(x = x, y = sigma)) +
  geom_line(linewidth = 1) +
  labs(
    x = "Concentration",
    y = "Response standard deviation"
  ) +
  theme_bw()
```

The response is still linear, but the scatter is not constant. A high-concentration point is expected to have a larger absolute residual than a low-concentration point. Treating those residuals as equally informative is the assumption that we will challenge in the next episode.

# Simulate OLS calibrations

```{r}
#| label: calibration-data

calibration_data <- lapply(
  designs,
  simulate_dataset,
  mean_fun = mean_fun,
  variance_fun = sigma_fun
)

models <- lapply(
  calibration_data,
  fit_ols
)
```

```{r}
#| label: calibration-plot
#| fig-cap: "The calibration can look linear even when variance clearly changes with concentration."

plot_design(
  calibration_data[["15 x 1"]],
  mean_fun
)
```

```{r}
#| label: residual-plot
#| fig-cap: "Residuals from OLS show the kind of pattern analysts should inspect."

plot_residuals(models[["15 x 1"]])
```

A residual plot is not decoration. It is a check of whether the model is describing the measurement process. If the residual spread grows with fitted response, the calibration line may still be linear, but the OLS uncertainty model is no longer appropriate.

# Predict unknown samples

```{r}
#| label: prediction

test_data <- generate_test_set(
  n = 1000,
  xmin = 1,
  xmax = 100,
  mean_fun = mean_fun,
  variance_fun = sigma_fun
)

prediction_tables <- lapply(
  models,
  predict_calibration_model,
  test_data = test_data,
  variance_fun = sigma_fun
)
```

```{r}
#| label: performance-table

performance_table <- data.table(
  design = names(prediction_tables),
  rmse = sapply(prediction_tables, metric_rmse),
  coverage = sapply(prediction_tables, metric_coverage),
  mean_interval_width = sapply(prediction_tables, metric_mean_pi_width),
  heteroscedasticity_detected = sapply(models, metric_hetero_detected)
)

kable(
  performance_table,
  digits = 3,
  caption = "OLS prediction performance when absolute variance changes with concentration."
)
```

The diagnostic test is intentionally simple, and with only 15 measurements it will not detect every heteroscedastic calibration. That is a practical lesson in itself. A small calibration experiment may not contain enough information to estimate a variance pattern reliably, even when that pattern is real.

# Take-home message

Linearity is not the whole calibration story. A method can have a straight mean response and still require a variance-aware analysis. When residual spread changes across the range, the next question is not whether the calibration is linear, but whether every point should have the same influence on the fitted line.
