---
title: "When variability changes: introducing weighted least squares"
author: "Andrea Bazzano"
format:
  html:
    toc: true
execute:
  echo: true
  warning: false
  message: false
---

# Why weights enter the story

When the response variability changes with concentration, ordinary least squares still fits a line, but it fits that line as if every point had the same response variance. Weighted least squares changes the question. Instead of treating all residuals equally, it gives more influence to observations expected to be more precise.

For approximately constant relative standard deviation, the response standard deviation is roughly proportional to concentration. The variance is then roughly proportional to concentration squared, which explains the common analytical rule to set weights as $w = 1/x^2$.

That rule is not magic. It is an assumption about the measurement process.

# Setup

```{r}
#| label: resources

library(here)
library(data.table)
library(ggplot2)
library(knitr)

source(here("R", "design_functions.R"))
source(here("R", "weight_functions.R"))
source(here("R", "fit_functions.R"))
source(here("R", "inversion_functions.R"))
source(here("R", "metrics_functions.R"))
source(here("R", "plotting_functions.R"))
source(here("R", "prediction_functions.R"))
source(here("R", "response_functions.R"))
source(here("R", "simulation_functions.R"))
source(here("R", "variance_functions.R"))
```

```{r}
#| label: setup

set.seed(123)

mean_fun <- f_linear
sigma_fun <- sigma_linear_constant_rsd

designs <- list(
  "15 x 1" = design_15x1(1, 100),
  "5 x 3" = design_5x3(1, 100),
  "3 x 5" = design_3x5(1, 100)
)
```

# Compare weight functions

Before fitting a model, this code puts all candidate weights on the same
concentration grid. The first curve is the inverse of the *simulated* response
variance. It is an oracle benchmark: in a real method we do not know it in
advance. The other curves are the practical shortcuts $1/x$ and $1/x^2$. Seeing
them together makes the modelling assumption visible: the steeper the curve,
the more strongly the fit prioritises low-concentration observations.

```{r}
#| label: weight-plot
#| fig-cap: "Different weighting rules encode different assumptions about variance."

weight_grid <- data.table(
  x = seq(1, 100, length.out = 200)
)

weight_plot_data <- rbindlist(
  list(
    data.table(
      x = weight_grid$x,
      method = "True inverse variance",
      weight = 1 / sigma_fun(weight_grid$x)^2
    ),
    data.table(
      x = weight_grid$x,
      method = "1/x",
      weight = weights_x(weight_grid$x)
    ),
    data.table(
      x = weight_grid$x,
      method = "1/x2",
      weight = weights_x2(weight_grid$x)
    )
  )
)

plot_weights(weight_plot_data)
```

A steeper weight curve gives much more importance to the low concentration standards. Under constant RSD, $\sigma(x) \propto x$, so inverse variance is proportional to $1/x^2$; the overlap in the figure is therefore the reason that rule works here. It becomes risky if the low end has a noise floor. The important question is therefore not whether a weighting rule is familiar, but whether it matches the method.

# Fit OLS and WLS models

```{r}
#| label: calibration-and-models

calibration_data <- lapply(
  designs,
  simulate_dataset,
  mean_fun = mean_fun,
  variance_fun = sigma_fun
)

models <- lapply(
  calibration_data,
  function(dt) {
    list(
      "OLS" = fit_ols(dt),
      "WLS true" = fit_wls_true(dt),
      "WLS estimated" = fit_wls_empirical(dt),
      "WLS 1/x" = fit_wls_x(dt),
      "WLS 1/x2" = fit_wls_x2(dt)
    )
  }
)
```

Empirical WLS requires replicated concentration levels. With 15 unreplicated levels, the calibration experiment has no direct way to estimate the variance at each level, so the empirical approach falls back to OLS. That is not a coding detail; it is an experimental design consequence.

The model list makes the comparison deliberately explicit. `WLS true` uses the
simulated standard deviations and is included only as a best-case benchmark.
`WLS estimated` first calculates the within-level replicate variance and then
uses its inverse as the regression weight. `WLS 1/x` and `WLS 1/x2` are fixed
rules that do not learn from the replicate data. Each fit is then inverted for
the same set of future samples, so differences in the table are differences in
the calibration strategy rather than differences in the samples being tested.

# Prediction performance

```{r}
#| label: prediction

test_data <- generate_test_set(
  n = 1000,
  xmin = 1,
  xmax = 100,
  mean_fun = mean_fun,
  variance_fun = sigma_fun
)

prediction_tables <- lapply(
  models,
  function(model_set) {
    lapply(
      model_set,
      predict_calibration_model,
      test_data = test_data,
      variance_fun = sigma_fun
    )
  }
)
```

```{r}
#| label: performance-table

performance_table <- rbindlist(
  lapply(
    names(prediction_tables),
    function(design_name) {
      data.table(
        design = design_name,
        model = names(prediction_tables[[design_name]]),
        rmse = sapply(prediction_tables[[design_name]], metric_rmse),
        bias = sapply(prediction_tables[[design_name]], metric_bias),
        coverage = sapply(prediction_tables[[design_name]], metric_coverage),
        mean_interval_width = sapply(prediction_tables[[design_name]], metric_mean_pi_width)
      )
    }
  )
)

kable(
  performance_table,
  digits = 3,
  caption = "Prediction performance under constant relative measurement error."
)
```

The true-variance WLS fit is the benchmark because it uses information that is available only in simulation. The simple $1/x^2$ rule should be close when the constant-RSD assumption is true. The $1/x$ rule is a milder correction, and empirical weighting depends strongly on whether the design contains enough replicates to estimate variability.

For a *validated linear method* with changing variance, I would choose **3 x 5
with estimated WLS** as the routine design. Five replicates at low, middle, and
high concentration let us estimate and check variability in the current run,
instead of simply assuming a familiar weight such as $1/x^2$. In this example
it also performs well for bias and prediction error. This is a routine choice,
not a replacement for validation: use more distributed levels during
validation when the aim is to establish linearity and the full variance pattern.

```{r}
#| label: error-profiles
#| fig-cap: "Relative prediction error by design and model."

for (design in names(prediction_tables)) {
  for (model in names(prediction_tables[[design]])) {
    print(
      plot_prediction_error(prediction_tables[[design]][[model]]) +
        labs(title = paste(design, model))
    )
  }
}
```

# Take-home message

Weighted least squares is useful because it connects calibration fitting with analytical precision. It is not a universal upgrade over OLS. If the weights describe the real variance structure, WLS can improve prediction and uncertainty statements. If the weights are only a habit, they can move the calibration in the wrong direction. For routine calibration of a validated linear method, 3 x 5 with data-estimated WLS is the practical choice here; use a broader, more distributed design when validating the response shape and variance function.
