---
title: "When the weighting model is only partly true"
author: "Calibration simulation series"
format:
  html:
    toc: true
execute:
  echo: true
  warning: false
  message: false
---

# The realistic complication

A clean constant-RSD model is useful for learning WLS, but real low-level analytical work often behaves differently. Near the LOQ, the response standard deviation may stop decreasing because blank noise, integration noise, or background correction dominates. Farther from the LOQ, relative error can become more stable.

This mixed behavior creates a problem for simple global weighting rules. A rule such as $1/x^2$ assumes that variance continues to shrink rapidly as concentration approaches zero. If the method has a low-concentration noise floor, that assumption can give too much influence to the lowest calibrators.

# Setup

```{r}
#| label: resources

library(here)
library(data.table)
library(ggplot2)
library(knitr)

source(here("R", "design_functions.R"))
source(here("R", "weight_functions.R"))
source(here("R", "fit_functions.R"))
source(here("R", "inversion_functions.R"))
source(here("R", "metrics_functions.R"))
source(here("R", "plotting_functions.R"))
source(here("R", "prediction_functions.R"))
source(here("R", "response_functions.R"))
source(here("R", "simulation_functions.R"))
source(here("R", "variance_functions.R"))
```

```{r}
#| label: setup

set.seed(321)

mean_fun <- f_linear
sigma_fun <- function(x) {
  sigma_loq_mixed(
    x,
    mean_fun = mean_fun,
    sigma_floor = 4,
    rsd = 0.07
  )
}

designs <- list(
  "15 x 1" = design_15x1(1, 100),
  "5 x 3" = design_5x3(1, 100),
  "3 x 5" = design_3x5(1, 100)
)
```

# The mixed variance pattern

```{r}
#| label: variance-pattern
#| fig-cap: "A mixed model: constant absolute error near the low end, then approximately constant RSD."

variance_grid <- data.table(
  x = seq(1, 100, length.out = 300)
)

variance_grid[, response := mean_fun(x)]
variance_grid[, sigma := sigma_fun(x)]
variance_grid[, rsd := sigma / response]

ggplot(variance_grid, aes(x = x, y = sigma)) +
  geom_line(linewidth = 1) +
  labs(
    x = "Concentration",
    y = "Response standard deviation"
  ) +
  theme_bw()

ggplot(variance_grid, aes(x = x, y = 100 * rsd)) +
  geom_line(linewidth = 1) +
  labs(
    x = "Concentration",
    y = "RSD (%)"
  ) +
  theme_bw()
```

The standard deviation has a floor at low concentration, while the RSD becomes more stable only after moving away from the low end. This is a familiar analytical situation: the method behaves differently near the quantification limit than it does in the comfortable middle of the range.

```{r}
#| label: weight-mismatch
#| fig-cap: "Simple global weighting rules do not fully match the mixed variance model."

weight_grid <- data.table(
  x = seq(1, 100, length.out = 300)
)

weight_plot_data <- rbindlist(
  list(
    data.table(
      x = weight_grid$x,
      method = "True mixed variance",
      weight = 1 / sigma_fun(weight_grid$x)^2
    ),
    data.table(
      x = weight_grid$x,
      method = "1/x",
      weight = weights_x(weight_grid$x)
    ),
    data.table(
      x = weight_grid$x,
      method = "1/x2",
      weight = weights_x2(weight_grid$x)
    )
  )
)

plot_weights(weight_plot_data)
```

The true weight curve flattens near the low end because precision does not keep improving indefinitely. The $1/x^2$ curve keeps rising. That mismatch is exactly where routine low-level calibration can become fragile.

# Fit and test models

```{r}
#| label: model-fit

calibration_data <- lapply(
  designs,
  simulate_dataset,
  mean_fun = mean_fun,
  variance_fun = sigma_fun
)

models <- lapply(
  calibration_data,
  function(dt) {
    list(
      "OLS" = fit_ols(dt),
      "WLS true" = fit_wls_true(dt),
      "WLS estimated" = fit_wls_empirical(dt),
      "WLS 1/x" = fit_wls_x(dt),
      "WLS 1/x2" = fit_wls_x2(dt)
    )
  }
)

test_data <- generate_test_set(
  n = 1000,
  xmin = 1,
  xmax = 100,
  mean_fun = mean_fun,
  variance_fun = sigma_fun
)

prediction_tables <- lapply(
  models,
  function(model_set) {
    lapply(
      model_set,
      predict_calibration_model,
      test_data = test_data,
      variance_fun = sigma_fun
    )
  }
)
```

```{r}
#| label: performance-table

performance_table <- rbindlist(
  lapply(
    names(prediction_tables),
    function(design_name) {
      data.table(
        design = design_name,
        model = names(prediction_tables[[design_name]]),
        rmse = sapply(prediction_tables[[design_name]], metric_rmse),
        coverage = sapply(prediction_tables[[design_name]], metric_coverage),
        mean_interval_width = sapply(prediction_tables[[design_name]], metric_mean_pi_width)
      )
    }
  )
)

kable(
  performance_table,
  digits = 3,
  caption = "Global prediction performance under a mixed variance model."
)
```

A global average can hide the region that matters most. For LOQ-related applications, the low concentration region deserves its own view.

```{r}
#| label: region-table

region_summary <- rbindlist(
  lapply(
    names(prediction_tables),
    function(design_name) {
      rbindlist(
        lapply(
          names(prediction_tables[[design_name]]),
          function(model_name) {
            dt <- copy(prediction_tables[[design_name]][[model_name]])
            dt[, region := cut(
              x_true,
              breaks = c(0, 10, 40, Inf),
              labels = c("Low", "Middle", "High")
            )]
            dt[, .(
              design = design_name,
              model = model_name,
              rmse = sqrt(mean(error^2)),
              mean_abs_relative_error = mean(abs(relative_error)),
              coverage = mean(x_true >= x_lpl & x_true <= x_upl)
            ), by = region]
          }
        )
      )
    }
  )
)

kable(
  region_summary,
  digits = 3,
  caption = "Performance by concentration region. Weight mismatch is most visible near the low end."
)
```

# Take-home message

WLS requires a variance model. A simple global rule may be convenient and may even be approximately right over part of the range, but that does not make it right near the LOQ. When the method has a low-concentration noise floor, the analyst should inspect low-level residuals, replicate behavior, and prediction performance separately from the global calibration summary.
