---
title: "When the weighting model is only partly true"
author: "Calibration simulation series"
format:
  html:
    toc: true
execute:
  echo: true
  warning: false
  message: false
---

# The realistic complication

A clean constant-RSD model is useful for learning WLS, but real low-level analytical work often behaves differently. Near the LOQ, the response standard deviation may stop decreasing because blank noise, integration noise, or background correction dominates. Farther from the LOQ, relative error can become more stable.

This mixed behavior creates a problem for simple global weighting rules. A rule such as $1/x^2$ assumes that variance continues to shrink rapidly as concentration approaches zero. If the method has a low-concentration noise floor, that assumption can give too much influence to the lowest calibrators.

# Setup

```{r}
#| label: resources

library(here)
library(data.table)
library(ggplot2)
library(knitr)

source(here("R", "design_functions.R"))
source(here("R", "weight_functions.R"))
source(here("R", "fit_functions.R"))
source(here("R", "inversion_functions.R"))
source(here("R", "metrics_functions.R"))
source(here("R", "plotting_functions.R"))
source(here("R", "prediction_functions.R"))
source(here("R", "response_functions.R"))
source(here("R", "simulation_functions.R"))
source(here("R", "variance_functions.R"))
```

```{r}
#| label: setup

set.seed(321)

mean_fun <- f_linear
sigma_fun <- sigma_linear_loq_mixed

designs <- list(
  "15 x 1" = design_15x1(1, 100),
  "5 x 3" = design_5x3(1, 100),
  "3 x 5" = design_3x5(1, 100)
)
```

# The mixed variance pattern

```{r}
#| label: variance-pattern
#| fig-cap: "A mixed model: constant absolute error near the low end, then approximately constant RSD."

variance_grid <- data.table(
  x = seq(1, 100, length.out = 300)
)

variance_grid[, response := mean_fun(x)]
variance_grid[, sigma := sigma_fun(x)]
variance_grid[, rsd := sigma / response]

ggplot(variance_grid, aes(x = x, y = sigma)) +
  geom_line(linewidth = 1) +
  labs(
    x = "Concentration",
    y = "Response standard deviation"
  ) +
  theme_bw()

ggplot(variance_grid, aes(x = x, y = 100 * rsd)) +
  geom_line(linewidth = 1) +
  labs(
    x = "Concentration",
    y = "RSD (%)"
  ) +
  theme_bw()
```

The standard deviation has a floor at low concentration, while the RSD becomes more stable only after moving away from the low end. This is a familiar analytical situation: the method behaves differently near the quantification limit than it does in the comfortable middle of the range.

```{r}
#| label: weight-mismatch
#| fig-cap: "Simple global weighting rules do not fully match the mixed variance model."

weight_grid <- data.table(
  x = seq(1, 100, length.out = 300)
)

weight_plot_data <- rbindlist(
  list(
    data.table(
      x = weight_grid$x,
      method = "True mixed variance",
      weight = 1 / sigma_fun(weight_grid$x)^2
    ),
    data.table(
      x = weight_grid$x,
      method = "1/x",
      weight = weights_x(weight_grid$x)
    ),
    data.table(
      x = weight_grid$x,
      method = "1/x2",
      weight = weights_x2(weight_grid$x)
    )
  )
)

plot_weights(weight_plot_data)
```

The true weight curve flattens near the low end because precision does not keep improving indefinitely. The $1/x^2$ curve keeps rising. That mismatch is exactly where routine low-level calibration can become fragile.

# Fit and test models

```{r}
#| label: model-fit

calibration_data <- lapply(
  designs,
  simulate_dataset,
  mean_fun = mean_fun,
  variance_fun = sigma_fun
)

models <- lapply(
  calibration_data,
  function(dt) {
    list(
      "OLS" = fit_ols(dt),
      "WLS true" = fit_wls_true(dt),
      "WLS estimated" = fit_wls_empirical(dt),
      "WLS 1/x" = fit_wls_x(dt),
      "WLS 1/x2" = fit_wls_x2(dt)
    )
  }
)

test_data <- generate_test_set(
  n = 1000,
  xmin = 1,
  xmax = 100,
  mean_fun = mean_fun,
  variance_fun = sigma_fun
)

prediction_tables <- lapply(
  models,
  function(model_set) {
    lapply(
      model_set,
      predict_calibration_model,
      test_data = test_data,
      variance_fun = sigma_fun
    )
  }
)
```

```{r}
#| label: performance-table

performance_table <- rbindlist(
  lapply(
    names(prediction_tables),
    function(design_name) {
      data.table(
        design = design_name,
        model = names(prediction_tables[[design_name]]),
        rmse = sapply(prediction_tables[[design_name]], metric_rmse),
        coverage = sapply(prediction_tables[[design_name]], metric_coverage),
        mean_interval_width = sapply(prediction_tables[[design_name]], metric_mean_pi_width)
      )
    }
  )
)

kable(
  performance_table,
  digits = 3,
  caption = "Global prediction performance under a mixed variance model."
)
```

A global average can hide the region that matters most. For LOQ-related applications, the low concentration region deserves its own view.

`coverage` is the proportion of simulated true concentrations that fall inside
their nominal 95% concentration prediction intervals. It is a calibration of
the uncertainty statement, not a measure of accuracy: a model can obtain high
coverage simply by making very wide intervals. For that reason it is reported
next to interval width, RMSE, and relative error. A useful interval should be
close to 95% coverage *and* no wider than necessary, especially in the region
where decisions are made.

```{r}
#| label: region-table

region_summary <- rbindlist(
  lapply(
    names(prediction_tables),
    function(design_name) {
      rbindlist(
        lapply(
          names(prediction_tables[[design_name]]),
          function(model_name) {
            dt <- copy(prediction_tables[[design_name]][[model_name]])
            dt[, region := cut(
              x_true,
              breaks = c(0, 10, 40, Inf),
              labels = c("Low", "Middle", "High")
            )]
            dt[, .(
              design = design_name,
              model = model_name,
              rmse = sqrt(mean(error^2)),
              mean_abs_relative_error = mean(abs(relative_error)),
              coverage = mean(x_true >= x_lpl & x_true <= x_upl)
            ), by = region]
          }
        )
      )
    }
  )
)

kable(
  region_summary,
  digits = 3,
  caption = "Performance by concentration region. Weight mismatch is most visible near the low end."
)
```

```{r}
#| label: coverage-profiles
#| fig-cap: "Empirical 95% prediction-interval coverage by concentration. The dashed line is the nominal target."

for (design in names(prediction_tables)) {
  for (model in names(prediction_tables[[design]])) {
    print(
      plot_prediction_interval_coverage(
        prediction_tables[[design]][[model]]
      ) +
        labs(title = paste(design, model))
    )
  }
}
```

The models can look very similar in overall RMSE and average prediction error.
That is not surprising once the setup is separated into its parts: all models
use the correct linear mean form, the mixed variance is only a moderate
departure from the simple rules, and global RMSE averages away the low-end
region where a mismatch matters most. Much of the remaining error is also
irreducible measurement noise. The regional table and coverage profiles are
therefore more informative than a small difference in one global summary.

# Take-home message

WLS requires a variance model. A simple global rule may be convenient and may even be approximately right over part of the range, but that does not make it right near the LOQ. When the method has a low-concentration noise floor, the analyst should inspect low-level residuals, replicate behavior, and prediction performance separately from the global calibration summary.

For this scenario, it would be reasonable to regard all the linear fits as
potentially suitable if their low-level recovery and interval coverage meet the
method's acceptance criteria. I would choose 3 x 5 with estimated WLS for
routine calibration *after validation*: its replicates let the laboratory
estimate and monitor the variance at low, middle, and high concentrations,
instead of assuming a fixed $1/x^2$ rule. This is statistically more auditable
than applying a familiar weight by default, but it does not replace a broader
validation design for testing linearity or discovering changes between the
three routine levels.
