---
title: "Can calibration designs detect small downward non-linearity equally well?"
author: "Calibration simulation series"
format:
  html:
    toc: true
execute:
  echo: true
  warning: false
  message: false
---

# The chemical calibration problem

In chemical calibration, non-linearity usually does not appear as a curve that bends upward forever. More often, the response begins to bend downward as the system approaches saturation, detector limitations, matrix effects, or the early part of a plateau.

That direction matters. A downward bend means that high-concentration samples produce less signal than a purely linear extrapolation from the low range would suggest. If the analyst forces a straight line through that response, concentration estimates can become biased, especially near the high end.

This episode asks whether the three 15-measurement designs detect small downward non-linearity equally well. The answer is expected to depend on both the number of distinct concentration levels and the measurement-error pattern.

# Setup

```{r}
#| label: resources

library(here)
library(data.table)
library(ggplot2)
library(knitr)

source(here("R", "design_functions.R"))
source(here("R", "weight_functions.R"))
source(here("R", "fit_functions.R"))
source(here("R", "inversion_functions.R"))
source(here("R", "metrics_functions.R"))
source(here("R", "plotting_functions.R"))
source(here("R", "prediction_functions.R"))
source(here("R", "response_functions.R"))
source(here("R", "simulation_functions.R"))
source(here("R", "variance_functions.R"))
```

```{r}
#| label: setup

set.seed(456)

mean_fun <- f_mild_downward_curvature
sigma_relative <- sigma_mild_downward_curvature_rsd

designs <- list(
  "15 x 1" = design_15x1(1, 100),
  "5 x 3" = design_5x3(1, 100),
  "3 x 5" = design_3x5(1, 100)
)
```

# The downward curvature

```{r}
#| label: curve-plot
#| fig-cap: "The simulated response bends downward, like the early stage of saturation."

curve_grid <- data.table(
  x = seq(1, 100, length.out = 300)
)

curve_grid[, y_curved := mean_fun(x)]
curve_grid[, y_linear_reference := x]
curve_grid[, difference_pct := 100 * (y_curved - y_linear_reference) / y_linear_reference]

ggplot(curve_grid, aes(x = x)) +
  geom_line(aes(y = y_curved), linewidth = 1) +
  geom_line(aes(y = y_linear_reference), linetype = 2) +
  labs(
    x = "Concentration",
    y = "Response"
  ) +
  theme_bw()

ggplot(curve_grid, aes(x = x, y = difference_pct)) +
  geom_line(linewidth = 1) +
  labs(
    x = "Concentration",
    y = "Deviation from linear response (%)"
  ) +
  theme_bw()
```

The dashed line is the ideal linear response. The solid line remains increasing, but its slope decreases with concentration. This is the small and clinically familiar danger zone: the calibration can still look almost linear, while the high end is already beginning to lose response.

# One noisy calibration

```{r}
#| label: example-calibration
#| fig-cap: "A single noisy calibration may not make mild downward curvature obvious."

example_data <- simulate_dataset(
  design_15x1(1, 100),
  mean_fun,
  sigma_homo
)

plot_design(example_data, mean_fun)
```

Visual inspection is useful, but it is not enough. A small downward bend can be hidden by normal analytical scatter, especially if the design has few distinct concentration levels.

# Repeated simulation

For each simulated calibration, we compare a linear model with a quadratic model. When response error is constant, we use an unweighted comparison. When relative error is constant, we use inverse-variance weights so that the curvature test matches the error structure.

```{r}
#| label: detection-simulation

n_sim <- 500

design_functions <- list(
  "15 x 1" = design_15x1,
  "5 x 3" = design_5x3,
  "3 x 5" = design_3x5
)

detection_results <- rbindlist(
  list(
    simulate_curvature_detection_power(
      design_functions = design_functions,
      mean_fun = mean_fun,
      variance_fun = sigma_homo,
      n_sim = n_sim
    )[, error_case := "Constant absolute error"],
    simulate_curvature_detection_power(
      design_functions = design_functions,
      mean_fun = mean_fun,
      variance_fun = sigma_relative,
      n_sim = n_sim,
      weights_fun = weights_from_simulated_sigma
    )[, error_case := "Constant relative error"]
  )
)

kable(
  detection_results,
  digits = 3,
  caption = "Estimated probability of detecting mild downward curvature."
)
```

```{r}
#| label: detection-plot
#| fig-cap: "None of the 15-measurement designs reaches 95% power to detect this deliberately mild curvature."

ggplot(
  detection_results,
  aes(
    x = design,
    y = detection_rate,
    fill = error_case
  )
) +
  geom_col(
    position = position_dodge(width = 0.8),
    width = 0.7
  ) +
  geom_hline(
    yintercept = 0.05,
    linetype = 2
  ) +
  geom_hline(
    yintercept = 0.95,
    linetype = 3
  ) +
  scale_y_continuous(
    limits = c(0, 1)
  ) +
  labs(
    x = "Calibration design",
    y = "Curvature detection rate",
    fill = "Error structure"
  ) +
  theme_bw()
```

The dashed line at 5% is the false-positive rate: even a truly linear response
will occasionally be labelled curved. It is **not** a target for detection. The
dotted line at 95% is a useful target for reliable detection. The bars show the
power of each design: the chance that it identifies the curvature when the
curvature is really present.

# Interpretation

For this deliberately *slight* curvature, **none of the 15-measurement designs
reaches 95% power**. Exact detection rates vary somewhat with the error
pattern, but all three are far below the level needed for reliable detection.
All three can occasionally find the bend, but none can reliably identify it.
Therefore, this simulation does not select one of the three as a
linearity-validation design. A non-significant curvature test should not be
read as proof that the response is linear.

Does that make curvature unimportant when prediction error is still suitable?
Not necessarily. The decision should be based on intended use. If bias,
prediction error, and interval coverage all meet predefined acceptance criteria
over the reportable range, a small bend may be operationally acceptable for
that range. It still matters because the bias can grow at the high end, the
response may change with time or matrix, and a linear model can become unsafe
outside the conditions that were checked. The goal of validation is therefore
not to detect every mathematically non-zero curve; it is to show that the
chosen model predicts well enough where analytical decisions are made.

For validation, use more distributed concentration levels plus replication so
that both shape and variance can be assessed against those acceptance criteria.
Once that work supports a linear model, 3 x 5 is a good compact routine design
for monitoring repeatability and variance at low, middle, and high levels.

# Take-home message

In chemical calibration, small non-linearity should usually be imagined as downward curvature or the beginning of saturation. With only 15 measurements, none of these designs reliably detects the deliberately mild bend simulated here. Choose the model on prediction performance, bias, coverage, and the intended range—not on a non-significant curvature test alone. The final episode turns that principle into a validation and routine-calibration workflow.
