---
title: "From validation to routine calibration: where should we invest our points?"
author: "Andrea Bazzano"
format:
  html:
    toc: true
execute:
  echo: true
  warning: false
  message: false
---

# One question, two phases

The series began with a simple question: **how should we invest our calibration
points?** The answer is not one design for every purpose. Validation and routine
calibration need different information.

During method validation, the laboratory is still learning whether the mean
response is linear and how its variance changes across the range. That requires
many well-spaced concentration levels and replication at each level. During
routine calibration, those questions should already have been answered. The
task changes to checking that the established model and variance structure
remain suitable for the current analytical run.

# The narrative across the six episodes

```{r}
#| label: resources

library(here)
library(data.table)
library(ggplot2)
library(knitr)

source(here("R", "design_functions.R"))
source(here("R", "weight_functions.R"))
source(here("R", "fit_functions.R"))
source(here("R", "inversion_functions.R"))
source(here("R", "metrics_functions.R"))
source(here("R", "plotting_functions.R"))
source(here("R", "prediction_functions.R"))
source(here("R", "response_functions.R"))
source(here("R", "simulation_functions.R"))
source(here("R", "variance_functions.R"))
```

```{r}
#| label: narrative-table

narrative <- data.table(
  context = c(
    "Linear response, constant absolute error",
    "Linear response, constant relative error",
    "Linear response, weights estimated from replicates",
    "Low-end noise floor plus relative error",
    "Slight downward curvature",
    "Validated routine method"
  ),
  lesson = c(
    "Spread points when the main aim is estimating the line.",
    "A straight mean response can still have changing variance.",
    "Weights are variance assumptions; replicates make them testable.",
    "Assess accuracy and interval coverage by concentration region.",
    "Few levels cannot establish that a subtle curve is absent.",
    "Monitor the validated model, variance, and prediction performance."
  )
)

kable(
  narrative,
  caption = "The same calibration-point budget answers different questions in different contexts."
)
```

# Validation design: learn the response and variance structure

A validation study should deliberately spend more than a routine calibration
run. Here is an illustrative design with seven distributed concentrations and
three independent replicates per level. It gives 21 measurements, not 15,
because validation is an investment in evidence rather than a routine-run
constraint. The concentration levels make curvature and regional departures
visible; the replicates make within-level variance estimable.

```{r}
#| label: validation-design
#| fig-cap: "Validation uses distributed levels and replication: enough positions to inspect shape, enough repeats to estimate variability."

validation_design <- data.table(
  x = rep(seq(1, 100, length.out = 7), each = 3),
  design = "Validation: 7 x 3"
)

routine_design <- copy(design_3x5(1, 100))
routine_design[, design := "Routine: 3 x 5"]

plot_design_comparison(
  rbindlist(list(validation_design, routine_design))
)
```

The exact number and placement of levels should be justified by the reportable
range, decision points, regulatory expectations, and prior method knowledge.
The key principle is not the literal 7 x 3 recipe. It is to include enough
levels to challenge linearity and enough replication to estimate the variance
function, including at the low end where a noise floor may appear.

# Routine design: monitor a model that has already been validated

For a validated linear method, 3 x 5 is a pragmatic routine design. Low,
middle, and high standards give a check across the reportable range, while five
replicates at each level allow an empirical estimate of local variability. An
estimated-WLS fit can then use information from the current run rather than a
fixed conventional weight.

```{r}
#| label: routine-example

set.seed(606)

mean_fun <- f_linear
sigma_fun <- sigma_linear_loq_mixed

routine_data <- simulate_dataset(
  routine_design,
  mean_fun = mean_fun,
  variance_fun = sigma_fun
)

routine_models <- list(
  "OLS" = fit_ols(routine_data),
  "WLS estimated" = fit_wls_empirical(routine_data),
  "WLS 1/x2" = fit_wls_x2(routine_data)
)
```

```{r}
#| label: routine-variance-check
#| fig-cap: "Replicates at the three routine levels make observed variability visible for each run."

observed_variance <- routine_data[
  ,
  .(
    response_sd = sd(y),
    response_mean = mean(y)
  ),
  by = x
]

ggplot(observed_variance, aes(x = x, y = response_sd)) +
  geom_point(size = 3) +
  geom_line() +
  labs(
    x = "Concentration",
    y = "Observed response SD"
  ) +
  theme_bw()
```

The three points do not re-validate the complete variance curve. They provide a
run-level check: do low, middle, and high variability look compatible with the
validated structure? If not, the laboratory should investigate the run or
return to a broader validation-style assessment rather than silently changing
weights.

# QC closes the loop

Calibration standards establish the model; quality-control (QC) samples test
whether it works on independent material. Place QCs at concentrations that
matter analytically—at least low, middle, and high, with particular attention
near decision limits and the LOQ. Track their bias, relative prediction error,
and prediction-interval coverage over time.

```{r}
#| label: qc-performance
#| fig-cap: "Routine QC checks prediction error at the concentrations where decisions are made."

qc_design <- data.table(
  x = rep(c(5, 50, 95), each = 100)
)

qc_data <- simulate_dataset(
  qc_design,
  mean_fun = mean_fun,
  variance_fun = sigma_fun
)

qc_predictions <- predict_calibration_model(
  routine_models[["WLS estimated"]],
  test_data = qc_data,
  variance_fun = sigma_fun
)

qc_summary <- qc_predictions[
  ,
  .(
    bias = mean(error),
    rmse = sqrt(mean(error^2)),
    mean_absolute_relative_error = mean(abs(relative_error))
  ),
  by = x_true
]

kable(
  qc_summary,
  digits = 3,
  caption = "Illustrative QC prediction checks by concentration level."
)
```

# Recommendation

Use a
multi-level, replicated design during validation to test linearity and define
the variance structure. Then use a 3 x 5 routine design with estimated WLS to
check that the validated variance structure remains stable and that QC
prediction performance is acceptable at low, middle, and high concentration.

This is statistically sound because WLS gains its efficiency from a credible
variance model, and routine replicates provide evidence for that model in the
current run. It is not a claim that 3 x 5 can independently prove linearity or
fully estimate an arbitrary variance curve. It is a compact monitoring design
that becomes defensible only after the broader validation work has established
the model.

# Take-home message

There is one coherent answer to the opening question. Invest calibration points
first in learning the method—shape and variance—then invest routine points in
verifying that the learned model still predicts acceptably. A 3 x 5,
data-estimated WLS calibration is a strong routine choice for a validated linear
method, provided QC results and replicate variability continue to support it.
